Complete Algebraic File Synchronization

### 2015-12-06 2015/paper.tex
Split paper into sections

\section{Definitions}
\input{def-filesystem}
\input{def-commands}

@@

% Ordering commands
% ----------------

\section{Ordering commands}\label{ordering}

We often encounter the case where we only have a set of commands without a specified order.
As we have seen above, this can occur after the first stage of update detection,
but, more importantly,
it is actually the task of the reconciler to determine whether there is an order,
and if yes, what order,
in which updates from multiple replicas can be applied to a filesystem.

In this section we describe an algorithm with which all possible orderings of a
simple set of commands can be generated, but before doing so, let us consider the following lemma.

\begin{mylem}\label{connected_changes}
Given a set of commands that
does not contain assertion commands,
and which can be applied to a filesystem in some order without breaking it,
and which contains commands on $n$ and $\nn$ where $\nn\descendant n$
but $\nn\neq\parent(n)$, then the set must also contain a commands
on each node between $n$ and $\nn$.
\end{mylem}
\begin{proof}
We prove that under the given conditions, the set must contain a command on $\parent(n)$.
Then, by repeating this statement, we how that the set must contain commands on every
ancestor of $n$ up to $\nn$.

Furthermore,
we prove this proposition for sequences, not sets, as if all sequences must contain a command on $\parent(n)$,
then so must sets because otherwise there would be no order in which they can be
applied to any filesystem.

Let $A$ be a sequence that satisfies the conditions;
we therefore know that it contains a command $\cxynv$ on $n$
and another command $\czwnnv$ on $\nn$.
We use an indirect proof and assume that there are no commands on $\parent(n)$ in $A$.
Next, we create a new sequence $A'\equiv A$ in which $\cxynv$ and $\czwnnv$ are next to each other.
If they are already next to each other in $A$, there is nothing to do.
Otherwise, consider the command next to $\cxynv$ in the direction where $\czwnnv$ is.
Let this command be $\cqrmv$.
If $\czwnnv$ is to the right, then $A$ looks like the following:
\[ A = \cdots\cc\cxynv\cc\cqrmv\cc\cdots\cc\czwnnv\cc\cdots \]
If $n\unrel m$, swap $\cxynv$ and $\cqrmv$. Based on \cref{ax_separate_commute} we know that the new
sequence is equivalent to $A$.
Otherwise, we know $m\neq\parent(n)$ as there are no commands on $\parent(n)$, and so
from \cref{ax_distantrel_breaks} we get $A\equiv\cbrk$ which contradicts our assumptions.
(Note that $A$ cannot contains assertion commands.)
By repeating this step we can therefore convert $A$ into $A'$ where $\cxynv$ and $\czwnnv$ are neighboring commands.
However, then \cref{ax_distantrel_breaks} applies to the sequence and therefore $A\equiv A'\equiv\cbrk$ which
is again a contradiction.
\end{proof}

We can now define our algorithm to generate orders of sets of commands.

Let $U$ be a simple set of commands
that we know can be applied to at least one filesystem in some order without breaking it.
Then we can split $U$ into a maximum number of disjunct subsets
(components) so that the nodes of two commands from two different subsets are unrelated.
These will simply be the components of the subset of the node forest that contains the nodes affected by the set.

From \cref{ax_separate_commute} we know that commands from different 
components can be applied in any order.
Therefore once we have ordered the commands inside a component, we generate
all permutations of the components themselves, including the ones
in which commands from separate components mix and overlap
(but where the possible orders determined for commands inside the components are respected).

Next, consider the ordering inside a component.

This is not a problem in the case of components containing a single command only.
(Note that $\cff$ can only appear in a component of one.)

In each component that is larger than one, 
we know that all nodes are related, and from \cref{connected_changes}
we also know that it must contain commands on all nodes in between.
Therefore the nodes
form a rooted ordered tree,
and the pairs of commands connected by the edges in the tree 
must be construction or destruction pairs.
(Otherwise, there would be no ordering of the component that is not equivalent to $\cbrk$
and via \cref{ax_separate_commute} there would be no ordering of the whole set, either.)

To order the commands in the component,
select the command with the topmost node, so that there would be no command
on its parent node (if there is one). 
If the command removes content at this node (it is of the form $\caa{X}{\ccharb}$), 
then it must follow all other commands in the component, and
all parent--child pairs in the whole component must be destruction pairs.
Otherwise, the command with the topmost node needs to precede the other commands,
and the component must be made of construction pairs.
These follow from the patterns of construction and destruction pairs, which
at no point can mix without causing any sequence to break all filesystems.

Then move the command with the topmost node to the beginning or the end of
the sequence we are creating out of the component,
and recursively start the whole algorithm again on the commands in the component that remain,
potentially splitting them into sub-components.
The result of this process is a set of possible orders for each component,
which are then combined into a set of possible orders for the whole set of commands.

\begin{mydef}{$\orderset{U}$.}
We use $\orderset{U}$ to denote the set of sequences that contain all commands in
the simple set $U$
and represent all orders generated by this algorithm.
\end{mydef}

It follows from the algorithm that:
\begin{mycor}\label{lemma:neighbor}
If $U$ contains commands on both $n$ and $\parent(n)$, then
there is a $U'\in\orderset{U}$ where they are next to each other.
\end{mycor}

As a specific case, we can regard a sequence of commands $S$
as a set and determine its $\orderset{S}$.
An important property of simple sequences is that
all of their valid orderings are equivalent,
and all equivalent sequences are mere reorderings:

\begin{mylem}\label{simple_reorder_equiv}
For any simple sequence $S\neq\cbrk$,
$\forall T: T\equiv S \Leftrightarrow T\in\orderset{S}$.
\end{mylem}
\begin{proof}
First, as $S$ only has at most one command on each node,
any sequence in $\orderset{S}$ either breaks a filesystem,
or it leaves it in the same state as $S$.

Second, based on the algorithm we also know that $\orderset{S}$ contains
all possible orderings of $S$ as the only ordering between commands
it prescribes are when the commands in the reverse order would break every filesystem.

If, for example, in a component or sub-component
made up of construction pairs,
the command on the topmost node, $\cxynv$, was not before all other commands
in the component as described by the algorithm,
then any resulting sequence would necessarily contain
a $\czwmv$ where $n\descendant m$, potentially followed by commands
on unrelated paths, and then followed by $\cxynv$.
From \cref{ax_separate_commute} we know we can convert this sequence
to an equivalent sequence by swapping the unrelated commands with
$\czmv$, which sequence would therefore have the commands on
$m$ and $n$ next to each other. Then from \cref{ax_distantrel_breaks,ax_directparent_breaks}
we know the sequence would break all filesystems
as a reversed construction pair is never a valid destruction pair.

We can prove that an order not generated by the algorithm
in case of components made of destruction pairs breaks all filesystems
in a similar way.

Also, from \cref{ax_separate_commute} we know
that the order between any other command pairs is not fixed in the sense
that if one order works, so will the other.

From these we know that all sequences in $\orderset{S}$
are equivalent, and that any sequence $T$ outside it
necessarily breaks all filesystems.

Therefore if $S\nequiv\cbrk$ then $S\in\orderset{S}$,
and $S$ cannot equivalent to any sequence of the same commands
not in $\orderset{S}$.
\end{proof}

It follows from this that given a simple set,
we can simply assume that it has an order.
Accordingly, we will at times treat not only sequences as sets,
but simple sets as sequences.

@@

If we start with two copies of the filesystem $\FS$,
and two different sequences are applied to the copies to yield $\FS_A:=A\FS$
and $\FS_B:=B\FS$, then our aim is to define sequences of commands $\reca$ and $\recb$
so that $\recb\FS_A$ and $\reca\FS_B$ would be as close to each other as possible.

We work based on the assumption that to achieve this, we need
to apply to $\FS_B$ the commands that have been applied to $\FS_A$, and \emph{vice versa}.
As some commands may have been applied to both filesystems, our first approximation
is $\reca = A\setminus B$ and $\recb = B\setminus A$
(in some suitable order).
This, however, will break both filesystems if there have been incompatible updates
in $A$ and $B$. 
Our aim is therefore to provide an algorithm that selects the commands 
$\reca \subset A\setminus B$
and $\recb \subset B\setminus A$ 
so that $\reca\FS_B\neq\fsbroken$ and $\recb\FS_A\neq\fsbroken$,
and show that these are the longest sequences with this property, that is,
adding any command left out from $\reca$ or $\recb$ will break the filesystems.

We assume that $A$ and $B$ are returned by an update detector,
and so they are simple sequences.
(Otherwise, they can be simplified to become simple sequences.)
We also assume that $\FS_A$ and $\FS_B$ are not broken.
With these assumptions,
we can now define independent sequences and
our reconciliation algorithm.

\begin{mydef}{Independent commands, sequences and sets.}
For two commands $\cxynv$ and $\czwmv$ we write $\cxynv\indep\czwmv$ if $\cxynv\cc\czwmv \equiv \czwnv\cc\cxynv \nequiv \cbrk$.
For two sequences or sets of commands $A$ and $B$ we write $A\indep B$ if
$\forall \alpha\in A \mbox{~and~} \forall \beta\in B: \alpha\indep\beta$.
\end{mydef}

\begin{mydef}{Reconciliation.}\label{def:reconciliation}
\begin{align*}
\reca = \{\alpha| & \\
& \alpha\in A\setminus B \\
& \wedge \\
& \forall \beta\in B\setminus A: \alpha\indep\beta\\
\} &
\end{align*}

In other words, the sequence $\reca$ we apply to $\FS_B$ 
is the largest subset of $A\setminus B$
that is independent of $B\setminus A$.
The order of commands in the sequence can be determined using the algorithm described in
\cref{ordering} as any subset of $A$ is also simple.
In a similar way, $\recb$ can be obtained by reversing $A$ and $B$.
\end{mydef}

We would now like to prove that $\reca$ and $\recb$ can be applied to the replicas,
that is, without loss of generality,
\[ \reca\FS_B\neq\fsbroken. \]
So that we can formalize statements needed to prove this,
let us introduce two relations that describe under what conditions
sequences of commands work, that is, do not break a filesystem.

@@

% Correctness of reconciliation
% -----------------------------

We are now ready to prove that the proposed algorithm for reconciliation is correct,
that is, applying its result is not going to break the replicas.
We reformulate the original proposition, $\reca\FS_B\neq\fsbroken$,
based on $\FS_B=B\FS$ and so we aim to prove that
$B\cc\reca$ works wherever $A$ and $B$ work:

\begin{myth}
If $A, B$ are simple, then $\worksc{B\cc \reca}{A,B}$,
where, to restate \cref{def:reconciliation},
% { % scope
% \renewcommand{\themydef}{\ref{def:reconciliation}}
% \begin{mydef}
\begin{align*}
\reca = \{\alpha| & \\
& \alpha\in A\setminus B \\
& \wedge \\
& \forall \beta\in B\setminus A: \alpha\indep\beta\\
\}. &
\end{align*}
% \end{mydef}
% } % scope
\end{myth}

This is trivial unless $\works{A,B}$, so we assume that it is true.
Let us first investigate the part of $A$ and $B$ that is excluded from
$\reca$: their intersection.

\begin{mylem}\label{can_move_intersection}
Let $A$ and $B$ be two simple sequences so that $\works{A, B}$.
Then their intersection, in some suitable order, also works on any filesystem
either of them works on:
$\worksc{A\cap B}{A}$, and, by symmetry,
$\worksc{A\cap B}{B}$.
\end{mylem}

\begin{proof}
The difficulty is that commands in $A\cap B$ can occur anywhere in $A$.
Therefore, first we prove the following:
if all commands are marked in $A$ that also appear in $B$,
then there is an $A'$ in $\orderset{A}$ in which all marked commands are at the beginning.

We show that $A$ can be transformed via equivalences so that this would be true.
Our proposition is that if the marked commands are not at the beginning, then
the sequence contains an unmarked command followed by a marked command.
We show that these can be swapped resulting in an equivalent sequence.
Then, by repeating this process similarly to bubble sorting, we can generate 
a suitable $A'$.

Let us consider therefore the marked command preceded by an unmarked command in $A$,
and let the marked command be $\cxynv$, 
and the preceding unmarked command be $\czwmv$.
As $A$ is simple and $\works{A}$, from 
\cref{ax_distantrel_breaks,ax_directchild_breaks,ax_directparent_breaks}
we know that these commands can only be on unrelated paths or form a construction or destruction pair.
In the first case, swapping the commands results in a sequence equivalent to $A$,
and we show that the last two cases are impossible as they would lead to contradiction.

In these cases, either $m=\parent(n)$ or $n=\parent(m)$.
If $B$ has a command on $m$, then
from \cref{lemma:neighbor}
we know that there is a $B'$ in $\orderset{B}$ where it is next to $\cxynv$ (also in $B$),
and from \cref{simple_reorder_equiv} we know that $B'\equiv B$.
When describing our ordering algorithm in \cref{ordering} we saw
that the command on the parent path deremines whether a pair of commands
is a construction or destruction pair,
and that this, in turn, determines whether the command on the child path must
precede or follow the other command as otherwise the sequence would break all filesystems.
This argument holds for both $A$ and $B$, and so the command on $m$ must be on
the same side of the command on $n$ in both sequences.

If $B$ has no command on $m$, then let $B'$ be $B$, 
but with an extra command added to it just before $\cxynv$
according to \cref{ax_child_assert,ax_parent_assert}, 
from which we know that $B'\equiv B$.
We also know that $B'$ is still minimal (but no longer simple).

In either case, therefore,
we have a minimal $B'$ which has a command on $m$ just before $\cxynv$ and $B'\equiv B$.
Let this command on $m$ be $\cqrmv$.
\begin{gather*}
A = \cdots\cc  \czwmv\cc  \cxynv\cc  \cdots \\
B' = \cdots\cc  \cqrmv\cc  \cxynv\cc  \cdots
\end{gather*}

As $\works{A,B}$ and so $\works{A,B'}$, 
from \cref{worksinputmatch}
we know that $Q=Z$.

Going back to the construction and destruction pairs we see that the output type of the first command
is always the same, and it is either $D$ or $\empt$ depending on the relationship between $n$ and $m$.
Therefore $R=W$ and $\cqrmv=\czwmv$
as $|\setd|=|\setb|=1$. 
If $B$ originally had a command on $m$,
this is a contradiction as $\czwmv$ was not marked, but we see it must also be in $B'$ and therefore in $B$.
If $B$ had no command on $m$,
this is a contradiction because the command on $m$ in $B'$ is an assertion command, so $Z=Q=R=W$, 
but $A$ contained no assertion commands.

We now know that there is an $A'\equiv A$ in which commands in $A\cap B$
are at the beginning, and therefore 
from \cref{worksextpostfix}
$\worksc{A\cap B}{A}$ and by symmetry $\worksc{A\cap B}{B}$.
\end{proof}

We would like to prove that $\worksc{B\cc \reca}{A,B}$.
From the above we know that this is equivalent to
\[ \worksc{\acb\cc \bma\cc \reca}{A,B}. \]
As we already know that $\reca\indep\bma$
and that $\worksc{\acb\cc \bma}{A,B}$
(as $\worksc{B}{A,B}$),
based on \cref{indep_prefix_combine}
we only need to prove that
\[ \worksc{\acb\cc \reca}{A,B}. \]
Using \cref{r_invmove} we can rephrase this as
\[ \worksc{\reca}{\acbi\cc A,\acbi\cc B}. \]

To restate some results above, we therefore already know the following things:
\begin{itemize}
\item $\amb$ and $\bma$ are simple as $A$ and $B$ are simple
\item $\amb \cap \bma = \emptyset$
\item $\reca$ is the largest subset of $\amb$ for which $\reca\indep\amb$
\item $\worksc{\amb, \bma}{\acbi\cc A,\acbi\cc B}$ 
because of \cref{r_invmove}
as $\worksc{A,B}{A,B}$,
and $A\equiv\acb\cc\amb$, and $B\equiv\acb\cc\bma$.
\end{itemize}

Our theorem is therefore proven if we prove the following lemma:
\newcommand{\condSimple}{(c1)}
\newcommand{\condDisj}{(c2)}
\newcommand{\condApr}{(c3)}
\newcommand{\condWork}{(c4)}
\begin{mylem}
If
   \begin{itemize}
   \item[\condSimple] $S$ and $T$ are simple sequences,
   \item[\condDisj] $S\cap T=\emptyset$,
   \item[\condApr] $S^*$ is the largest subset of $S$ where $S^*\indep T$, and
   \item[\condWork] $\worksc{S,T}{\seqset{C}}$ for some set of sequences $\seqset{C}$,
   \end{itemize}
then
\[ \worksc{S^*}{\seqset{C}}. \]
\end{mylem}

\begin{proof}
This is trivial unless $\works{\seqset{C}}$, so we will assume that it is true
and therefore $\works{S,T}$.

The proof is similar to that of \cref{can_move_intersection}.
We mark all commands in $S$ that are in its subset $S^*$, and
we prove that $S$ can be transformed into $S'$ via equivalences
so that all marked commands would be at its beginning.
If so, then based on \cref{worksextpostfix}
$\worksc{S^*}{\seqset{C}}$ as $\worksc{S'}{\seqset{C}}$ and $S'=S^*\cc S_0$ 
where $S_0$ contains the remaining (conflicting) commands.

Again we know that if $S$ does not already have all marked commands at its beginning,
then there is an unmarked command followed by a marked one.
We show that these commands are independent and so they can be swapped
resulting in an equivalent sequence.
By repeating this process we can generate a suitable $S'$.

Let therefore the marked command in $S$ be $\cxynv$
and the preceding unmarked command be $\czwmv$.
As $S$ is simple and $\works{S}$, from 
\cref{ax_distantrel_breaks,ax_directchild_breaks,ax_directparent_breaks}
we know that these commands can only be on unrelated paths or form a construction or destruction pair.
In the first case, swapping the commands results in a sequence equivalent to $S$;
we only need to show that the other two cases are not possible as they would lead to a contradiction.

In the last two cases, we know that either $m=\parent(n)$ or $n=\parent(m)$.
We also know that because of {\condApr} there must be 
a command $\cqrov$ in $T$ which is not independent of $\czwmv$
as $\czwmv$ is not part of $S^*$.
\begin{gather*}
S = \cdots\cc  \czwmv\cc  \cxynv\cc  \cdots \\
T = \cdots\cc  \cqrov\cc \cdots
\end{gather*}
We know none of these commands is an assertion command, and 
from {\condDisj} that $\cqrov\neq\czwmv$.
Therefore, based on \cref{unrelated_is_independent} we know this means that
either $m\descendantEq o$ or $o\descendantEq m$.
From {\condApr} we also know that $\cxynv\indep \cqrov$,
and so because of \cref{unrelated_is_independent},
$n\unrel o$.

We therefore have four cases considering the relationships between $n,m$ and $o$:
\begin{itemize}
\item $n=\parent(m) \wedge o\descendantEq m$.
   This would mean that $o\descendantEq n$ or $n=\parent(o)$, which contradicts $n\unrel o$.
\item $n=\parent(m) \wedge m\descendantEq o$.
   This would mean that $n\descendant o$, which contradicts $n\unrel o$.
\item $m=\parent(n) \wedge o\descendantEq m$.
   This would mean that $o\descendant n$, which contradicts $n\unrel o$.
\item $m=\parent(n) \wedge m\descendantEq o$.
   Let us continue with this case.
\end{itemize}

We know that $T$ cannot have a command on $m$ as $m=\parent(n)$ and $\cxynv\indep T$,
and so $m\neq o$ and therefore $m\descendant o$.
This means we can create a new sequence $T'\equiv T$ by inserting a command on 
$m$ into $T$ before $\cqrov$
according to \cref{ax_parent_assert}.
We now have $\cxynv\indep T\equiv T'$ which is a contradiction as
$T'$ has a command on $m$ and $m=\parent(n)$.
\end{proof}

\subsection{Reconciliation is complete}

% Reconciliation is maximal
% -------------------------
% This is where we're using |D|=1
% TODO C.f. strict / liberal etc. reconciliation in references

The reconciliation algorithm defined above is also complete, that is,
it is not possible to apply any further commands from $A\setminus B$ to $\FS_B$.
To show this, we are going to prove the following \namecref{rec_is_complete}:

\begin{myth}\label{rec_is_complete}
If $A$ and $B$ are simple sequences,
and we select a sequence $S\subset \amb$ so that
it contains a command $\cxynv$ for which $\cxynv\nindep\bma$,
then $B\cc S$ either breaks all filesystems $A$ and $B$ do not break
(that is, $S$ breaks all possible $\FS_B$ replicas),
or it overrides incompatible changes introduced in the replica by $B$.
\end{myth}
Overriding changes is not desirable as in these cases the reconciler
does not have enough information to decide whether to promote
the change introduced by $A$ over that introduced by $B$ or vice versa.
These cases should be treated as conflicts and
the affected nodes or commands marked for review
by a human or a different system.
\begin{proof}
The proof is similar to the ones we have seen above.
Without loss of generality, we assume that $\cxynv$ is the first command in $S$
that is not independent of $\bma$.
If so, we can split $S$ into $S_0\cc\cxynv\cc S_1$ where $S_0\indep\bma$.
Let us also split $\bma$ into $T_0\cc\czwmv\cc T_1$ where $T_1\indep\cxynv$.
From \cref{can_move_intersection} we know that
$B\cc S \equiv \acb\cc\bma\cc S$.
We also know commands in $S_0$ and $\bma$ commute, and so this is equivalent to
$\acb\cc S_0\cc\bma\cc\cxynv\cc S_1$.
Expanding $\bma$ and swapping $T_1$ and $\cxynv$ we get
\[ \acb\cc S_0\cc T_0 \cc \czwmv \cc \cxynv\cc T_1\cc S_1. \]

First, we prove that this sequence breaks all filesystems
unless $n=m$.
Let us therefore suppose $n\neq m$ and, to use an inverse proof,
that the sequence works on $\FS$ where $A\FS\neq\fsbroken$.
If so, its prefix, $\acb\cc S_0\cc T_0 \cc \czwmv \cc \cxynv$ must also work.

We know $\cxynv\nindep\czwmv$, and so from \cref{unrelated_is_independent}, $n\nunrel m$.
As the sequence works, $\czwmv\cc\cxynv$ can only be a construction or destruction pair.
As $\czwmv$ and $T_0$ originate from $B\setminus A$ and $B$ is simple, we also know there are
no commands on $m$ in $A\cap B$ or $T_0$,
and that there are no commands on $m$ in $S_0$ as $S_0\indep\bma$.

On the one hand we therefore know that $\FS(m)$ must be of type $Z$,
but on the other, from \cref{ax_directchild_breaks,ax_directparent_breaks},
we also know that $\cxynv$ cannot be applied to $\FS$ without
changing $\FS(m)$ first.
As $A\FS\neq\fsbroken$, this means that $A$ must contain a command on $m$.
Let this command be $\cqrmv$.

Clearly $Q=Z$ (from \cref{worksinputmatch}), and from the construction and destruction
pairs we also see that $R=W$ must hold, where $W=\cchard$ or $W=\ccharb$.
As $|\setd|=1$ and $|\setb|=1$, this also means that $v_R=v_W$ and therefore
$\cqrmv = \czwmv$.
This is a contradiction as $\cqrmv\in A$, but $\czwmv\in\bma$.

The only possibility is therefore $n=m$.
As $A$ and $B$ are simple, we therefore know that these are the first commands on $n$,
and so (again from \cref{worksinputmatch}) $Z=X$.
We also know $\valvw\neq\valvy$ as otherwise the two commands would be equal and
would be included in $A\cap B$,
but this means that $\cxynv$, from $A$, overrides a change introduced 
by $\czwnv$, from $B$, which should be treated as a conflict.
\end{proof}

From the proof it is apparent that this result depends on $|\setd|=1$.

>>>>

\section{Definitions}

\input{def-filesystem}

\input{def-commands}

@@

\input{ordering}

@@

\input{rec-intro}

@@

\input{rec-sound}

\input{rec-complete}
